Connecting R with D3 for dynamic graphics, to explore multivariate data with tours

### 2019-01-21 paper-tourrd3.Rmd
Some more details on technical aspects.

The data format expected by D3 is in JSON format, which combines two basic programming paradigms: a collection of name/value pairs, and an ordered list of values. R's preferred data formats include data frames, vectors and matrices. Every time a new projection has been calculated with the tour path, the resulting matrix needs to be converted to JSON and sent to D3. The code to send the D3 data looks like this:

```
session$sendCustomMessage(type = "data", message = toJSON(j))
```

This code is from the observe environment from the `server.R` file. It converts the matrix of projected data points to JSON format, and sends it to JavaScript with the id data. When parsed in D3 by its `data()` method, it is converted back into a logical 2D array where the columns are queried first, then the rows. If column names are included in the JSON, the column indices are strings; otherwise they are integers starting from 1.
All of the code required to render the scatterplots and legends, along with colours, is JavaScript code in the file `d3anim.js`. In particular, the data from R is handled with the following code:

@@

The projections are calculated using the tour object in an `observe()` environment, which repeatedly runs the code until a reactive variable is changed, at which point it is "invalidated" and then re-started. The projections are calculated using the following code
block:

@@

>>>>

To set up the interface for the app, we need to load the relevant scripts into the Shiny app and assign a section for the resulting plots. This is done when setting up the user interface. We import D3 and our plotting code via the `tags$script` (for web links) and `includeScript` (for reading from a full path). We use `tags$div` to assign an id for the output section that can be accessed in the D3 code.

```
tags$script(src = "https://d3js.org/d3.v4.min.js"),
includeScript(system.file("js/d3anim.js", package = "tourrGUID3")),
tags$div(id = "d3_output")
```

The data format expected by D3 is in JSON format, which combines two basic programming paradigms: a collection of name/value pairs, and an ordered list of values. R's preferred data formats include data frames, vectors and matrices. Every time a new projection has been calculated with the tour path, the resulting matrix needs to be converted to JSON and sent to D3. Using a named list we can send multiple JSON datasets to D3, e.g. to draw both the data points (stored in dataframe d) and the projection axes (stored in dataframe a). Converting dataframes will pass the column names to JSON. The code to send the D3 data looks like this:

```
session$sendCustomMessage(type = "data", message = list(d = toJSON(d), a = toJSON(a)))
```

This code is from the observe environment from the `server.R` file. It converts the matrix of projected data points to JSON format, and sends it to JavaScript with the id data. The list entries of the "message" can parsed in D3 by its `data()` method, e.g. `data(message.d)` to access the projected data points, and we can access each column through the column names assigned in the original dataframe, and loop over all rows for rendering.
All of the code required to render the scatterplots and legends, along with colours, is JavaScript code in the file `d3anim.js`. In particular, the data from R is handled with the following code:

@@

The projections are calculated using the tour object in an `observe()` environment, which re-executes the code whenever it is invalidated. The invalidation is either by a change in reactive value inside the code block, or we can schedule a re-execution by explicitly invalidating the observer after a selected interval. The projections are calculated using the following code block:

@@

## Troubleshooting

Fixing bugs in the JavaScript code can be cumbersome, as R and Shiny will not report any errors. Tracing JavaScript errors can be done when using the JavaScript console in the web browser. For example, in Google Chrome the console can be accessed via the "Developer Tools" option found under "Moore Tools" in the control menu. Typical errors that we encountered were version dependent syntax in D3, e.g. for axis definitions or scaling.

### 2019-01-21 paper-tourrd3.tex
Some more details on technical aspects.

\abstract{%
The tourr package in R has several algorithms and displays for showing
multivariate data as a sequence of low-dimensional projections. It can
display as a movie but has no capacity for interaction, such as stop/go,
change tour type, drop/add variables. The tourrGui package provides
these sorts of controls, but the interface is programmed with the dated
RGtk2 package. This work explores using custom messages to pass data
from R to D3 for viewing, using the Shiny framework.
}

\subsection{Introduction}\label{introduction}

The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995};
Cook et al. \protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of
systematically generating and displaying projections of high-dimensional
spaces in order for the viewer to examine the multivariate distribution
of data. It can do this either randomly, or by picking projections
judged interesting according to some criterion or index function. The
tourr package (Wickham et al. \protect\hyperlink{ref-tourr}{2011})
provides the computing and display in R (R Core Team
\protect\hyperlink{ref-R}{2018}; Ihaka and Gentleman
\protect\hyperlink{ref-ihaka:1996}{1996}) to make several types of
tours: grand, guided, little and local. The projection dimension can be
chosen between one and the number of variables in the data. The display,
though, has no capacity for interaction. The viewer can watch the tour
like a movie, but not pause it and restart, or change tour type, or
number of variables.

@@

\subsection{Creating a tour, with the tourr
package}\label{creating-a-tour-with-the-tourr-package}

@@

A tour path is a smooth sequence of projection matrices, \(p\times d\),
that when combined with a matrix of n data points, \(n\times p\), and a
rendering method, produces a steady stream of \(d\)-dimensional views of
the data. Each tour is initialised with the \texttt{new\_tour()} method,
which instantiates a tour object and takes as arguments the data \(X\),
the tour method, e.g. \texttt{guided\_tour()}, and the starting basis.
Once initialised, a new target plane is chosen, and a series of steps
along a geodesic path from starting to target plane are generated by
interpolation.

@@

\subsection{\texorpdfstring{Connecting the tour projections to D3
display using
\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}

@@

The data format expected by D3 is in JSON format, which combines two
basic programming paradigms: a collection of name/value pairs, and an
ordered list of values. R's preferred data formats include data frames,
vectors and matrices. Every time a new projection has been calculated
with the tour path, the resulting matrix needs to be converted to JSON
and sent to D3. The code to send the D3 data looks like this:

\begin{verbatim}
session$sendCustomMessage(type = "data", message = toJSON(j))
\end{verbatim}

This code is from the observe environment from the \texttt{server.R}
file. It converts the matrix of projected data points to JSON format,
and sends it to JavaScript with the id data. When parsed in D3 by its
\texttt{data()} method, it is converted back into a logical 2D array
where the columns are queried first, then the rows. If column names are
included in the JSON, the column indices are strings; otherwise they are
integers starting from 1. All of the code required to render the
scatterplots and legends, along with colours, is JavaScript code in the
file \texttt{d3anim.js}. In particular, the data from R is handled with
the following code:

@@

\subsection{Getting projections}\label{getting-projections}

@@

The projections are calculated using the tour object in an
\texttt{observe()} environment, which repeatedly runs the code until a
reactive variable is changed, at which point it is ``invalidated'' and
then re-started. The projections are calculated using the following code
block:

@@

\subsection{Pros and cons}\label{pros-and-cons}

@@

\subsection{Summary}\label{summary}

@@

\subsection{Code}\label{code}

@@

\subsection{Acknowledgements}\label{acknowledgements}

@@

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-D3}{}
Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
Data-Driven Documents.'' \emph{IEEE Transactions on Visualization and
Computer Graphics} 17 (12): 2301--9.
\url{http://vis.stanford.edu/papers/d3}.

\hypertarget{ref-shiny}{}
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
McPherson. 2017. \emph{Shiny: Web Application Framework for R}.
\url{https://CRAN.R-project.org/package=shiny}.

\hypertarget{ref-gt_pp}{}
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (4): 155--72.

\hypertarget{ref-gt_pp_mc}{}
Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
``Grand Tours, Projection Pursuit Guided Tours and Manual Controls.''

\hypertarget{ref-tourrGui}{}
Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
gWidgets Gui for the Tour to Explore High-Dimensional Data Using
Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
(6): 1--12.

\hypertarget{ref-ihaka:1996}{}
Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
Analysis and Graphics.'' \emph{Journal of Computational and Graphical
Statistics} 5 (3): 299--314.

\hypertarget{ref-RGtk2}{}
Lawrence, Michael, and Duncan Temple Lang. 2010. ``RGtk2: A Graphical
User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
37 (8): 1--52. \url{http://www.jstatsoft.org/v37/i08/}.

\hypertarget{ref-R}{}
R Core Team. 2018. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

\hypertarget{ref-tourr}{}
Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
``Tourr: An R Package for Exploring Multivariate Data with
Projections.'' \emph{Journal of Statistical Software} 40 (2): 1--18.

>>>>

\abstract{%
The tourr package in R has several algorithms and displays for showing
multivariate data as a sequence of low-dimensional projections. It can
display as a movie but has no capacity for interaction, such as stop/go,
change tour type, drop/add variables. The tourrGui package provides
these sorts of controls, but the interface is programmed with the dated
RGtk2 package. This work explores using custom messages to pass data
from R to D3 for viewing, using the Shiny framework.\\
}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995},
\protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of systematically
generating and displaying projections of high-dimensional spaces in
order for the viewer to examine the multivariate distribution of data.
It can do this either randomly, or by picking projections judged
interesting according to some criterion or index function. The tourr
package (Wickham et al. \protect\hyperlink{ref-tourr}{2011}) provides
the computing and display in R (R Core Team
\protect\hyperlink{ref-R}{2018}; Ihaka and Gentleman
\protect\hyperlink{ref-ihaka:1996}{1996}) to make several types of
tours: grand, guided, little and local. The projection dimension can be
chosen between one and the number of variables in the data. The display,
though, has no capacity for interaction. The viewer can watch the tour
like a movie, but not pause it and restart, or change tour type, or
number of variables.

@@

\hypertarget{creating-a-tour-with-the-tourr-package}{%
\subsection{Creating a tour, with the tourr
package}\label{creating-a-tour-with-the-tourr-package}}

@@

A tour path is a smooth sequence of projection matrices, \(p\times d\),
that when combined with a matrix of n data points, \(n\times p\), and a
rendering method, produces a steady stream of \(d\)-dimensional views of
the data. Each tour is initialised with the \texttt{new\_tour()} method,
which instantiates a tour object and takes as arguments the data \(X\),
the tour method, e.g.~\texttt{guided\_tour()}, and the starting basis.
Once initialised, a new target plane is chosen, and a series of steps
along a geodesic path from starting to target plane are generated by
interpolation.

@@

\hypertarget{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}{%
\subsection{\texorpdfstring{Connecting the tour projections to D3
display using
\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}}

@@

To set up the interface for the app, we need to load the relevant
scripts into the Shiny app and assign a section for the resulting plots.
This is done when setting up the user interface. We import D3 and our
plotting code via the \texttt{tags\$script} (for web links) and
\texttt{includeScript} (for reading from a full path). We use
\texttt{tags\$div} to assign an id for the output section that can be
accessed in the D3 code.

\begin{verbatim}
tags$script(src = "https://d3js.org/d3.v4.min.js"),
includeScript(system.file("js/d3anim.js", package = "tourrGUID3")),
tags$div(id = "d3_output")
\end{verbatim}

The data format expected by D3 is in JSON format, which combines two
basic programming paradigms: a collection of name/value pairs, and an
ordered list of values. R's preferred data formats include data frames,
vectors and matrices. Every time a new projection has been calculated
with the tour path, the resulting matrix needs to be converted to JSON
and sent to D3. Using a named list we can send multiple JSON datasets to
D3, e.g.~to draw both the data points (stored in dataframe d) and the
projection axes (stored in dataframe a). Converting dataframes will pass
the column names to JSON. The code to send the D3 data looks like this:

\begin{verbatim}
session$sendCustomMessage(type = "data", message = list(d = toJSON(d), a = toJSON(a)))
\end{verbatim}

This code is from the observe environment from the \texttt{server.R}
file. It converts the matrix of projected data points to JSON format,
and sends it to JavaScript with the id data. The list entries of the
``message'' can parsed in D3 by its \texttt{data()} method,
e.g.~\texttt{data(message.d)} to access the projected data points, and
we can access each column through the column names assigned in the
original dataframe, and loop over all rows for rendering. All of the
code required to render the scatterplots and legends, along with
colours, is JavaScript code in the file \texttt{d3anim.js}. In
particular, the data from R is handled with the following code:

@@

\hypertarget{getting-projections}{%
\subsection{Getting projections}\label{getting-projections}}

@@

The projections are calculated using the tour object in an
\texttt{observe()} environment, which re-executes the code whenever it
is invalidated. The invalidation is either by a change in reactive value
inside the code block, or we can schedule a re-execution by explicitly
invalidating the observer after a selected interval. The projections are
calculated using the following code block:

@@

\hypertarget{troubleshooting}{%
\subsection{Troubleshooting}\label{troubleshooting}}

Fixing bugs in the JavaScript code can be cumbersome, as R and Shiny
will not report any errors. Tracing JavaScript errors can be done when
using the JavaScript console in the web browser. For example, in Google
Chrome the console can be accessed via the ``Developer Tools'' option
found under ``Moore Tools'' in the control menu. Typical errors that we
encountered were version dependent syntax in D3, e.g.~for axis
definitions or scaling.

\hypertarget{pros-and-cons}{%
\subsection{Pros and cons}\label{pros-and-cons}}

@@

\hypertarget{summary}{%
\subsection{Summary}\label{summary}}

@@

\hypertarget{code}{%
\subsection{Code}\label{code}}

@@

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

@@

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-D3}{}%
Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
Data-Driven Documents.'' \emph{IEEE Transactions on Visualization and
Computer Graphics} 17 (12): 2301--9.
\url{http://vis.stanford.edu/papers/d3}.

\leavevmode\hypertarget{ref-shiny}{}%
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
McPherson. 2017. \emph{Shiny: Web Application Framework for R}.
\url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\hypertarget{ref-gt_pp}{}%
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (4): 155--72.

\leavevmode\hypertarget{ref-gt_pp_mc}{}%
Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
``Grand Tours, Projection Pursuit Guided Tours and Manual Controls.''

\leavevmode\hypertarget{ref-tourrGui}{}%
Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
gWidgets Gui for the Tour to Explore High-Dimensional Data Using
Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
(6): 1--12.

\leavevmode\hypertarget{ref-ihaka:1996}{}%
Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
Analysis and Graphics.'' \emph{Journal of Computational and Graphical
Statistics} 5 (3): 299--314.

\leavevmode\hypertarget{ref-RGtk2}{}%
Lawrence, Michael, and Duncan Temple Lang. 2010. ``RGtk2: A Graphical
User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
37 (8): 1--52. \url{http://www.jstatsoft.org/v37/i08/}.

\leavevmode\hypertarget{ref-R}{}%
R Core Team. 2018. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

\leavevmode\hypertarget{ref-tourr}{}%
Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
``Tourr: An R Package for Exploring Multivariate Data with
Projections.'' \emph{Journal of Statistical Software} 40 (2): 1--18.

\bibliography{references.bib}
